Geometric signs reveal changes in social structures and networks during the Western Eurasian Aurignacian

### 2025-10-28 paper/paper.qmd
fix typos spotted by Aleah

We hypothesize that these phases, and the environmental changes across the them, may correspond to changes in dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and distribution? What do these patterns tell us about dynamics of social organization throughout the Aurignacian? To answer these questions we use seriation analysis, network analysis, and PerMANOVA tests.

@@

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`). This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type number and diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to very strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.

@@

Our results are consistent with @Banks_dErrico_Zilhao_2013 who used eco-cultural niche modeling to show that the Heinrich Stadial 4 (HS4) at the very beginning of the Early Aurignacian resulted in a expansion of people into new environmental niches, which they hypothesize was associated with an expansion of social networks both within and between populations. This correlates with our results which show a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased sign type correlation with geographic distance. Similarly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. This geographic expansion may explain our results, where we find weaker, more dispersed connections in the Evolved Aurignacian.

Our results do not suggest that geometric signs serve as markers of a structured ethnic identity <!-- BM: be more precise about what our findings do not support --> , as found with Aurignacian personal ornaments in @Vanhaeren_d. However, the data used by @Vanhaeren_d was also analyzed by @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015 <!-- BM: expand a bit more on what this study did and found -->to find that cultural identity by-descent with modification <!-- BM: expand this to define to show readers we know what this is --> and isolation-by-distance is a more likely explanation for the distribution of Aurignacian personal ornaments. Our results support the pattern of cultural identity by-descent with modification (although they do not support isolation-by-distance [@tbl-mantel-table]), as well as the patterns of signs and other artwork as a form of cultural in-group and out-group communication found by @Sauvet_2018 and @baker2024evidence. <!-- BM: want to engage specifically with Baker et al. maybe do this earlier in this para, start the para with reference specifically to Baker et al.  -->

@@

PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017; @Shennan_Crema_Kerig_2015]. The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value. The pseudo-F statistic compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups. Larger pseudo-F statistics indicate more separation between groups. The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The perMANOVA was performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.

>>>>

We hypothesize that these phases, and the environmental changes across them, may correspond to changes in dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and distribution? What do these patterns tell us about dynamics of social organization throughout the Aurignacian? To answer these questions we use seriation analysis, network analysis, and PerMANOVA tests.

@@

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`). This confirms our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type number and diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to very strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The perMANOVA tests show that the changes between the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.

@@

Our results are consistent with @Banks_dErrico_Zilhao_2013 who used eco-cultural niche modeling to show that the Heinrich Stadial 4 (HS4) at the very beginning of the Early Aurignacian resulted in an expansion of people into new environmental niches, which they hypothesize was associated with an expansion of social networks both within and between populations. This correlates with our results which show a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased sign type correlation with geographic distance. Similarly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. This geographic expansion may explain our results, where we find weaker, more dispersed connections in the Evolved Aurignacian.

Our results support a general theory of geometric signs as a form of cultural identification and communication during the Palaeolithic, as proposed by previous work [e.g. @Sauvet_2018 and @baker2024evidence]. However, our results do not include geographically distinct cultural groups, and so our findings contrast with previous work such as @Vanhaeren_d and @baker2024evidence who found distinct cultural clusters in their analysis of personal ornaments. One possible explanation of this contrast is that a distinguishing factor between the Aurignacian and Gravettian is the appearance of geographically distinct cultural groups in the Gravettian, as reported by @baker2024evidence. We might hypothesize that the distribution found in the Evolved Aurignacian [@fig-group-map] evolved into the clusters proposed by @baker2024evidence for the Gravettian period.

That said, our results also contrast with @Vanhaeren_d, who found nine ethnically structured groups using evidence from the Aurignacian period. One possible explanation for this is that the geometric signs studied here encode different kinds of group information compared to the personal ornament objects studied by @Vanhaeren_d. On one hand, we hypothesize that geometric signs are not a useful proxy for detecting geographically distinct cultural groups in the Palaeolithic, compared to personal ornaments. On the other hand, the data used by @Vanhaeren_d was also analyzed by @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015, who used Approximate Bayesian Computation to simulate and compare Aurignacian personal ornament distribution as a result of structured ethnic identity [e.g. ] versus as a result of cultural identity by-descent with modification and isolation-by-distance. Their results show the latter was best supported by the data. Cultural identity by-descent with modification is defined as group difference generated by unconscious cultural mutation and drift, and largely influenced by spatial proximity to other groups. These results from  @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015 complicate claims by @Vanhaeren_d for geographically distinct cultural groups in the Aurignacian, and validate our findings.

While our results do not entirely support the findings of @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015, given that our Mantel tests did not find a statistically significant correlation between geographic distance and sign type similarity (@tbl-mantel-table), cultural identity by-descent with modification seems a likely explanation for some of the patterns seen in our results.

@@

PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017; @Shennan_Crema_Kerig_2015]. The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value. The pseudo-F statistic compares the total sum of squared dissimilarities between groups to the total sum of squared dissimilarities within groups. Larger pseudo-F statistics indicate more separation between groups. The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The perMANOVA was performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
